Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-06-24 joint_inf.tex
Minor changes and suggestions.

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$.
For the other term, we define
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
as a mapping of ancestral states and tip states to ancestral state splits.
It is defined by if the tip states have their complements taken or not: if we exclude a set of tip labels $\alignmentColumn$ from the set of site patterns by taking its complement, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise it is simply $\ancestralStateColumn$.
Note $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective for each $\alignmentColumn_i$.
We define
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
By considering the case requiring complementing and that not requiring complementing separately,
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
By the surjectivity of $\ancestralToSplit(\alignmentColumn_i, \cdot)$, we also have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

@@

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with branch lengths}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2,3\}\}$&
    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)^2   (1-w)(1+y)^2\\
                    (1-x)^2   (1+w)(1+y)^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2,3\}\}$&
                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,2,3\}$&
                $\emptyset$&
                $(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x)(1-x)(1+w)(1+y)^2\\
                        (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1+x)^2   (1-w)(1+y)^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                    (1-x)^2   (1+w)(1+y)^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1-x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1-w)(1-y)^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state partitions.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Appendix for full calculations.}
\label{tab:likelihoods}
\end{table}

\subsubsection{Upper bound for Farris likelihood}

@@

The full Farris log likelihood takes one of three values---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
The log likelihood is

@@

Seeking an upper bound, we use the fact that, for $x',y'\in[0,1]$,
(1-x'^2+y'^2-x'^2y'^2) = (1-x'^2)(1+y'^2) = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y),
with similar substitutions for $(1+x'^2-y'^2-x'^2y'^2)$ and $(1-x'^2-y'^2+x'^2y'^2)$.
Doing so results in

@@

\subsubsection{Lower bound for Felsenstein partial likelihood}

>>>>

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$.
For the other term, we define
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
as a mapping of ancestral states and tip states to ancestral state splits.
It is defined by if the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is not contained in $\siteSplitSet$ and thus we must take a complement to become a site split, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise it is simply $\ancestralStateColumn$.
Note $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective for each $\alignmentColumn_i$.
We define
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
By considering the case requiring complementing and that not requiring complementing separately,
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
By the surjectivity of $\ancestralToSplit(\alignmentColumn_i, \cdot)$, we also have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

@@

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
%EM Is it indicated anywhere how the internal nodes are numbered? I would assume that 1 would be connected to leaf 1. If that's true please indicate here.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

\begin{table}
%EM I wonder why you use primes on all variables here but not in tab:likelihoods.
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2,3\}\}$&
    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    %EM Shouldn't this be \{1\}\\ ? Or perhaps the ordering of internal nodes is not what I thought?
                    \{2\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)^2   (1-w)(1+y)^2\\
                    (1-x)^2   (1+w)(1+y)^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2,3\}\}$&
                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,2,3\}$&
                $\emptyset$&
                $(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x)(1-x)(1+w)(1+y)^2\\
                        (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1+x)^2   (1-w)(1+y)^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                    (1-x)^2   (1+w)(1+y)^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1-x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1-w)(1-y)^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state partitions.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Appendix for full calculations.}
\label{tab:likelihoods}
\end{table}

\subsubsection{Upper bound for Farris tree likelihood}

@@

The full Farris tree log likelihood takes one of three values depending on branch lengths---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is

@@

Seeking an upper bound, we use the facts that, for $x',y'\in[0,1]$,
(1-x'^2+y'^2-x'^2y'^2) & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
(1+x'^2-y'^2-x'^2y'^2) & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y').
Factoring and making these substitutions results in

@@

\subsubsection{Lower bound for Felsenstein tree partial likelihood}
